---
title: "Deep mutational scanning microcins toxins"
subtitle: "Part 2, Differential Expression Analysis with DESeq2"
author: "Luiz Carlos Vieira"
date: "08/07/2024"
---


Differential expression analysis with DESeq2 involves multiple steps. First, DESeq2 will model the raw 
Counts, using normalization factors (size factors) to account for differences in library depth. 

Second, it will estimate the gene-wise dispersion and shrink these estimates to generate more accurate
estimates of dispersion to model the Counts. 

Third, DESeq2 will fit the negative binomial model and perform hypothesis testing using the Wald test or 
Likelihood Ratio Test.

[Love, M.I., 2014](https://doi.org/10.1186/s13059-014-0550-8)


## Libraies
```{r message=FALSE, include=FALSE}
library(DESeq2)
library(ggplot2)
library(ggrepel)
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
library(openxlsx)

#set current directory as working directory

setwd(getwd())
```



# Loading count matrix
```{r}
#count_matrix <- read.csv(paste(getwd(), "/results/read_", sep=""), header=TRUE, sep = ",")
count_matrix <- read.csv("results/read_counts_MccV_full_v02.csv")

head(count_matrix)
```


## dataframe Counts as matrix
```{r}
# filling NAs with zeros.
count_matrix[is.na(count_matrix)] <- 0

# making ID the rownames of the matrix
row.names(count_matrix) <- count_matrix$ID

# selectiong columns from count matrix
cols <- c("CountsLib2Glu1", "CountsLib2Glu2", "CountsLib2Glu3", "CountsLib2Ara1", "CountsLib2Ara2", "CountsLib2Ara3")

Counts <- count_matrix[cols]
Counts <- as.matrix(Counts)
```


## Creating a coldata table

info as sample_id and conditions.
```{r}
coldata <- data.frame(row.names = colnames(Counts),
                       group= rep(c("Glue", "Ara"), each=3))



# convert column "condition" to "factor"
coldata$group <- factor(coldata$group,
                        levels = c("Glue", "Ara"))

coldata
```


# DESeq2 object
```{r}
dds <- DESeqDataSetFromMatrix(countData=Counts, colData=coldata, design= ~group)

dds
```


## Filters out row sums smaller than 100 Counts.
```{r}
filtro <- rowSums(counts(dds)) >= 100
 
dds <- dds[filtro, ]
```


### relevel() to define reference level. 
```{r}
dds$group <- relevel(dds$group, ref = "Glue")
```


## running deseq2
```{r}
ddsDE <- DESeq(dds)

ddsDE
```

### Checking group comparions:
```{r}
resultsNames(ddsDE)
```


### Getting the results 

Setting a alpha value of 0.05 for each comparison
```{r}
# setting contrast: 1 -factor of interest, 2 - numerator (treated) e 3- denominator (control).
res <- results(ddsDE, alpha = 0.05, contrast=c("group", "Ara", "Glue"))
res
```


## Summary results
```{r}
summary(res)
```

From these results we can see that more than 30744 peptides had a lfc < 0.


### Results column description
```{r}
mcols(res)$description
```



## Adding significance DE column to the df results

True DE results if padj<0.05 and absolute log2FoldChange > 1
```{r}
# group one vs zero
res$DE <- ifelse(res$padj<0.05, "True", "False")
res[which(abs(res$log2FoldChange)<1),'DE'] <- "False"
res <- res[order(res$padj),]
res

```



# saving res as a data frame
```{r}
res_df <- as.data.frame(res)
res_df
```



# filter peptides diferential expressed with lgf <= -1
```{r}
peps_DE <- dplyr::filter(res_df, log2FoldChange <= (-1) & padj <0.05)
peps_DE
```

# Dataframe with all results from deseq2


```{r}
res_df$ID <- rownames(res_df)

res_full <- left_join(count_matrix, res_df, by= "ID")
res_full 

```


```{r}
write.csv(res_full, 'results/res_full_deseq2.csv')
```



### -------------------------------------------------------------------------###
#                       Quality control DESeq2 Analysis
###------------------------------------------------------------------------- ###


## sizeFactors

sizeFactors are related to the library size (total number of reads in the library), but are calculated in 
such a way that compensate the effect that most RNAseq samples, the most higly expressed genes take up the
majority of the reads.

```{r, fig.height=3, fig.width=8}
par(mfrow=c(1,3))

## Check the size factors
barplot(sizeFactors(ddsDE),  main =('Size Factors'))

## Total number of raw counts per sample
barplot(colSums(counts(dds)), main =('Raw counts'))

## Total number of normalized counts per sample
barplot(colSums(counts(ddsDE, normalized=T)), main =('Normalized Counts'))
```

NOTE: Calculate gene-specific normalization factors (size factors) accounts for of technical biases such as 
differing dependence on GC content, gene length.



## Regularized log transformation for PCA plots and heatmaps 
```{r}
# Regularized log transformation 
rld <- rlog(ddsDE)

rld_mtx <- assay(rld)
```



### Histogram of reads counts distribuition.
```{r, fig.height=5, fig.width=8}
hist(rld_mtx, xlab = 'Normalized ReadsCounts', main= 'rlog Transformad Data' )
```


## Reads counts distribuition after transformation
```{r, fig.height=5, fig.width=8}
x <- rld_mtx
corRainbow = rainbow(dim(x)[2])

plot(density(x[,1]), col = corRainbow[1], lwd=2,
     xlab="Normalized ReadsCounts", ylab="Density", main= "Normalized Readscounts Distribuition",
     ylim=c(0, max(density(x[,1])$y)+.02 ) )
  
for( i in 2:dim(x)[2] )
lines(density(x[,i]), col=corRainbow[i], lwd=2)
legend("topright", cex=1.1, colnames(x), lty=rep(1,dim(x)[2]), col=corRainbow)
```



## Estimate gene-wise dispersion

```{r, fig.height=5, fig.width=8}    
#jpeg(filename ="results/Estimate_Dispersion.jpeg", height=600, width=800, quality = 200, res = 120)
plotDispEsts(ddsDE, main="Estimate Dispersion")
```
DESeq2 uses a specific measure of dispersion (α) related to the mean (μ) and variance of the data:
    
    Var = μ + α*μ^2.

So 0.01 is equal to 10% variation around the mean.


Red line represents the estimate for the expected dispersion value for genes of a given expression strength.

A good representation of data is the data points scattered around the curve, with the dispersion decreasing 
with increasing mean expression levels.


# PCA 

Plot PCA by column variable
```{r, fig.height=5, fig.width=8}
#jpeg(filename ="results/PCA_plot.jpeg", height=600, width=800, pointsize=12, res = 120)
plotPCA(rld, intgroup = "group", ntop = 500) +
  theme_bw() +
  geom_point(size = 5) +
  ggtitle(label = "Principal Component Analysis (PCA)")
```


```{r, fig.height=3, fig.width=8}
pca_data = rld_mtx[,c(1, 2, 3, 4, 5, 6)]

# Transpose the data
transposed_data <- t(pca_data)

# Run PCA
pca_result <- prcomp(transposed_data, center= TRUE, scale.= FALSE)

#results
df <- as.data.frame(pca_result$x)

# creating the factors
df$Group <- factor(c(rep("Uninduced", 3), rep("Induced", 3)))

ggplot(df, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 4) +
  theme_light() +
  ggtitle("Principal Component Analysis (PCA)") +
  xlab(paste("PC1: ", round(summary(pca_result)$importance[2, 1]*100, 2), "% variance")) +
  ylab(paste("PC2: ", round(summary(pca_result)$importance[2, 2]*100, 2), "% variance"))
  #scale_x_continuous(breaks= seq(-15, 15, by = 5))

#ggsave('results/pca_induced_x_uninduced_reads_counts.jpeg', width=8 , height=3 )

```




## Sample Similarity
```{r, fig.height=5, fig.width=8}
#jpeg(filename="results/similarity_plot.jpeg", height=1000, width=1000, quality = 200, res = 120)
sampleDists <- dist(t(rld_mtx))

sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- rld$group
colnames(sampleDistMatrix) <- rld$group

corBlues <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)

pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=corBlues)

```



## Distribution of adjusted p-values
```{r, fig.height=3, fig.width=9}
par(mfrow=c(1,3))
hist(res$padj, breaks=50, col="gray", main =('Analyzing values of ajusted p-values'),
     xlab = 'p-value ajustado',
     ylab = 'Frequency')


```
High frequency around zero, represents DEs results


## MA-Plot of LFCs vs Normalized countReads

```{r, fig.height=4, fig.width=9}
#jpeg("results/plotMA-dispersions2.jpeg", width = 1200, height = 700, quality = 200, res = 120)

#par(mfrow=c(1,3))
plotMA(res, main='Ara vs Glue', 
       xlab='Mean counts')

```

Each point is a peptide and blue points are differential expressed genes.

Point above the line are up-regulated peptides and below the line down-regulated peptides

Obs: The triangular points, represents points out of the plot window. 



# - - - - - - - - - - - - - - - - -  - - - - - - - - - - - - - - - - - - - - #
#             Visualization of Differencial Expressed Peptides 
# - - - - - - - - - - - - - - - -  - - - - - - - - - - - - - - - - - - - - - #


## Volcano plot 1uM IPTG vs 0uM IPTG
```{r}
topDE = filter(res_df, abs(log2FoldChange) >= 3)
topDE
```


```{r, fig.height=5, fig.width=8}
legen <- res_df[row.names(res_df) %in% rownames(topDE), ]

volcano = ggplot(res_df, aes(log2FoldChange, -log10(pvalue))) + 
  geom_point(aes(col=DE)) +
  scale_color_manual(values=c("red", "blue"))


volcano + geom_label_repel(data= legen, aes(label= row.names(legen)), size=3, box.padding = unit(2, "lines"), point.padding = unit(5, "points"), max.overlaps = 100) +
   labs(title= "Volcano plot of Diferencial Expressed Peptides",
       subtitle = "Comparison of Ara vs Glue",
       x= "log2 fold change", 
       y= "p-value (-log10)",
       color="Differential Expressed") +
  scale_x_continuous(limits = c(-20, 20)) +
  theme_bw() + coord_cartesian(clip = "off")

ggsave(filename = "results/volcano_plot_one.jpeg", width = 10, height = 6, dpi = 600)
```



### Volcano plot 10uM IPTG vs 0uM IPTG
```{r, fig.height=5, fig.width=8}
legen <- ten[rownames(ten) %in% rownames(topDE), ]

volcano = ggplot(ten, aes(log2FoldChange, -log10(pvalue))) + 
  geom_point(aes(col=DE)) +
  scale_color_manual(values=c("red", "blue"))


volcano +
  geom_label_repel(data= legen, aes(label= row.names(legen)), size=3, box.padding = unit(3, "lines"),
                   point.padding = unit(5, "points"), max.overlaps =100) +
   labs(title= "Volcano plot of Diferencial Expressed Peptides",
       subtitle = "Comparison of 10uM vs 0uM",
       x= "log2 fold change", 
       y= "p-value (-log10)",
       color="Differential Expressed") +
  scale_x_continuous(limits = c(-20, 20)) +
  theme_bw() + coord_cartesian(clip = "off")

ggsave(filename = "results/volcano_plot_ten.jpeg", width = 8, height = 5, dpi = 300)
```



## Volcano plot 100uM IPTG vs 0uM IPTG
```{r, fig.height=5, fig.width=8}
legen <- hnd[rownames(hnd) %in% rownames(topDE), ]


volcano = ggplot(hnd, aes(log2FoldChange, -log10(pvalue))) + 
  geom_point(aes(col=DE)) +
  scale_color_manual(values=c("red", "blue"))


volcano + 
  geom_label_repel(data= legen, aes(label= row.names(legen)), size=3, box.padding = unit(5, "lines"),
                   point.padding = unit(5, "points"), max.overlaps =150) +
  
  labs(title= "Volcano plot of Diferencial Expressed Peptides",
       subtitle = "Comparison of 100uM vs 0uM",
       x= "log2 fold change", 
       y= "p-value (-log10)",
       color="Differential Expressed") +
  scale_x_continuous(limits = c(-10, 8)) +
  theme_bw() + coord_cartesian(clip = "off")

ggsave(file="results/volcano_plot_hdn.jpeg", height=5, width=8, dpi=300)
```



### Clustering genes with heatmap
```{r, fig.height=10, fig.width=9}
jpeg("results/pheatmap_z-score_top50DE_v01.jpeg", height=2000, width=1000, res = 120)

#color pallet
cores_heat <- colorRampPalette(brewer.pal(9, "YlOrRd"))(250)
#cores_heat <- colorRampPalette(brewer.pal(9, "Blues"))(250)

# order matrix by difference in mean variation
tx <- order(-rowVars(rld_mtx))
mat <- rld_mtx[tx, ]

# z-score
mat <- (mat - rowMeans(mat)) / sd(mat)

# Filter matrix by peptides of interest
mat <- rld_mtx[row.names(rld_mtx) %in% rownames(topDE), ]

# description df
descr <- data.frame(row.names = colnames(Counts), 
                    Group = as.factor(rep(c("0uM", "1uM", "10uM", "100um"), each=3)))
descr$Group <- factor(descr$Group, levels = c("0uM", "1uM", "10uM", "100um"))

#Heatmap
pheatmap(head(mat, 50), scale="row", cluster_rows=TRUE, show_rownames=TRUE,
          show_colnames=T, cluster_cols=F, annotation_col=descr, col=cores_heat,
         main = "The top 50 Peptides Differencial Expressed, colored by z-score")


```



# -------------------------------------------------------------------------------------------#
#                                 Download results tables
# -------------------------------------------------------------------------------------------#


### Results DESeq2
```{r}
#write.xlsx(df_full, 'results/res_full_deseq2.xlsx', colNames = TRUE, rowNames = F)
```



# --------------------------------------------------------------------------------------------#
```{r}
sessionInfo()
```

